---
title: "demo_2"
author: "Theo Killian"
date: "July 1, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Practical: 03: Core approaches in Bioconductor

"My favorite gene is found on what chromasome?"

### Core infrastructure

S4 Objects. Why? Interoperability. Robustness. What? Interface (what you want to
know about) and implementation (what you don’t want to know about!)

```{r library}
library(Biostrings)
```

```{r}
head(methods(class = "DNAStringSet") )
```

### Biostrings

DNA, amino acid, and other biological sequences. See earlier example in 01:
Introdction to R and Bioconductor.

### GRanges

GRanges(): genomic coordinates to represent annotations (exons, genes,
regulatory marks, ...) and data (called peaks, variants, aligned reads)

GRangesList(): genomic coordinates grouped into list elements (e.g., paired-end
reads; exons grouped by transcript)

```{r}
library(GenomicRanges)
```

```{r}
exon <- GRanges(c("chr1:20-30:+", "chr1:40-50:+", "chr1:45-55:+"))
# three ranges on chromasome 1 (don't necessarily have to be chromasomes, could
# be contigs) and says that it's on the "+" strand
exon
```

```{r}
start(exon)
```

```{r}
end(exon)
```

```{r}
width(exon)
```

```{r}
disjoin(exon)
```

```{r}
# create SNP
snp <- GRanges(c("chr1:12345", "chr1:23456"))
snp

```

```{r}
countOverlaps(snp, exon)
```

```{r}
countOverlaps(exon, snp)
```

```{r}
snp %over% exon
```

```{r}
snp[snp %over% exon]
```

```{r}
promoters(exon)
```

```{r}
cvg <- coverage(exon)
# returns compact representation of "Rle" run length encoding
# coverage(exon) %>% unlist() %>% as.integer()
```

```{r}
# as(cvg, GRanges)
# cvg
grcvg <- cvg %>% as("GRanges")
```

```{r}
mcols(grcvg)
```

```{r}
df <- DataFrame(i = 1:3, dna = DNAStringSet(c("AAA", "CCC", "GGG")), gr = exon,
                green = "Recycle!")
```


```{r}
gr <- GRanges("A", IRanges(c(10, 20, 22), width=5), "+")
shift(gr, 1)                            # intra-range
```

```{r}
range(gr)                               # inter-range
```

```{r}
reduce(gr)                              # inter-range
```

```{r}
snps <- GRanges("A", IRanges(c(11, 17, 24), width=1))
findOverlaps(snps, gr)                  # between-range
```

```{r}
setdiff(range(gr), gr)                  # 'introns'
```

### SummarizedExperiment

```{r}
## always use BioManager::install("package")
## because it looks on CRAN and Bioconductor

library(SummarizedExperiment)
library(airway)
library(ggplot2)
library(dplyr)
library(rtracklayer)
```

```{r}
data(airway)
airway
```

```{r}
colData(airway)
```

```{r}
dim(assay(airway))
# head(assay(airway))
```

```{r}
airway[, airway$dex %in% "trt"]
```

```{r}
chr14 <- as(seqinfo(rowRanges(airway)), "GRanges")["14"]
airway[airway %over% chr14,]
```

```{r}
colSums(assay(airway))
```

```{r}
rowRanges(airway)
```

```{r}
metadata(airway)
```

### The matrix-like behavior of SummarizedExperiment

```{r}
set.seed(123)
m <- matrix(rnorm(12), nrow = 4, ncol = 3,
    dimnames = list(letters[1:4], LETTERS[1:3]))
m
```

```{r}
dim(m)
```

```{r}
m[1:2, 2:1]
```

```{r}
m[1:2,]
```

```{r}
m[1,]
```

```{r}
m[,1]
```

```{r}
m[ 1, , drop = FALSE]
```

### Basic summaries of assay() values

```{r}
m
m+1
```

```{r}
# or applying a transformation
abs(m + 1)
```

```{r}
colSums(assay(airway))
# pretty sure that these are gene counts per sample
```

```{r}
rowMeans(assay(airway))
# average of the rows is sort of like the mean expression of the gene
```

```{r}
hist(rowMeans(assay(airway)))
```

### Subsetting SummarizedExperiment

```{r}
airway$dex
# pick out untreated samples
```

```{r}
untreated <- airway$dex == "untrt"
untreated_subset <- airway[,untreated]
untreated_subset$dex
## better
untrt <- airway[, airway$dex == "untrt"]
ridx <- rowSums(assay(airway)) > 0
head(ridx)
```

```{r}
sum(ridx/length(ridx))
```

```{r}
table(ridx)
```

```{r}
nonzero_airway <- airway[ridx, ]
```

```{r}
hist(rowMeans(log(assay(nonzero_airway) + 1)))
```

### The list-like interface of GRangesList

```{r}
l <- list(a = 1:5, b = month.abb)
```

```{r}
names(l)
```

```{r}
length(l)
```

```{r}
l[c(2, 1)]
```

```{r}
l[2] # list of length 1, containing element 2 of original list
```

```{r}
l[[2]]     # element 2 of original list
# double square bracket subsetting is now outside of the list, now it's a vector 
```

```{r}
ridx <- rowMeans(x) > 0
m[ridx, , drop=FALSE]
```

```{r}
# remove assay rows that are all zero
ridx <- rowSums(assay(airway)) > 0

# get non-zero rows
se <- airway[ridx,]

# transform into pseudolog
ave_log_expr <- rowMeans(log(1 + assay(se)))

# transform into vector
tbl <- tibble::enframe(ave_log_expr, "gene", "ave_log_expr")

# NOTE: enframe()
# Converting vectors to data frames, and vice versa!!!
ggplot(tbl, aes(ave_log_expr)) + geom_density()
```

```{r}
r <- rowRanges(airway)
r
```

```{r}
seqinfo(r)
```

```{r}
seqinfo(r) %>% as("GRanges") %>% subset(seqnames == 14)
```

```{r}
rowRanges(airway) %over% chr14
```

## From files to Bioconductor objects

```{r}

```

```{r}
data(cpgIslands, package="Gviz")
cpgIslands$name <- letters[1:10]
cpgIslands$score <- sample(10)
cpgIslands
```

```{r}
bed_file <- tempfile(fileext = ".bed")
basename(bed_file)
```

```{r}
export(cpgIslands, bed_file)
cat(readLines(bed_file), sep = "\n")
```

```{r}
import(bed_file)
```

```{r}
gtf <- system.file(package = "airway", "extdata",
                   "Homo_sapiens.GRCh37.75_subset.gtf")
import(gtf)
```

### FASTQ files

```{r}

```

### The GenomicAlignments package

```{r}
library(GenomicAlignments)
```

```{r}
sessionInfo()
```

## 05: Bioconductor Annotation Resources

```{r}
# https://uclouvain-cbio.github.io/BSS2019/05_Annotations.html
BiocManager::available("^org\\.")
```

```{r}
library(AnnotationHub)
query(AnnotationHub(), "^org\\.")
```

```{r}
library(org.Hs.eg.db)
```

```{r}
head(keys(org.Hs.eg.db))
```

```{r}
eid <- sample(keys(org.Hs.eg.db), 10)
```

```{r}
mapIds(org.Hs.eg.db, eid, "SYMBOL", "ENTREZID")
```

```{r}
map <- select(org.Hs.eg.db, eid, c("SYMBOL", "GO"), "ENTREZID")
```

```{r}
dim(map)
```

```{r}
head(map)
```

### GO.db

```{r}
library(GO.db)
```

### Transcript annotations

TxDb.Hsapiens.UCSC.hg38.knownGene

TxDb packages contain information about gene models (exon, gene, transcript
coordinates). There are a number of TxDb packages available to install

```{r}
library(dplyr)    # for `%>%`
```

```{r}
BiocManager::available("^TxDb") %>% tibble::enframe(name = NULL)
```

and to download from AnnotationHub...

```{r}
query(AnnotationHub(), "^TxDb\\.")
```

```{r}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

### exons(), transcripts(), genes()

```{r}
exons(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
ex <- exons(TxDb.Hsapiens.UCSC.hg38.knownGene, columns = "GENEID")
ex
```

```{r}
std_ex <- keepStandardChromosomes(ex, pruning.mode="coarse")
std_ex
```

```{r}
table(seqnames(std_ex))
```

```{r}
std_ex[width(std_ex) > 10000]
```

### ensembldb

```{r}
library(ensembldb)
```

```{r}
query(AnnotationHub(), c("^EnsDb\\.", "Ensembl 96"))
```

### Accessing online resources (biomart)

```{r}
library(biomaRt)
head(listMarts(), 3)                      ## list marts
head(listDatasets(useMart("ensembl")), 3) ## mart datasets
ensembl <-                                ## fully specified mart
    useMart("ensembl", dataset = "hsapiens_gene_ensembl")

head(listFilters(ensembl), 3)             ## filters
myFilter <- "chromosome_name"
substr(filterOptions(myFilter, ensembl), 1, 50) ## return values
myValues <- c("21", "22")
head(listAttributes(ensembl), 3)          ## attributes
myAttributes <- c("ensembl_gene_id","chromosome_name")

## assemble and query the mart
res <- getBM(attributes =  myAttributes, filters =  myFilter,
             values =  myValues, mart = ensembl)
```

### KEGGREST

```{r}
library(KEGGREST)
```

### AnnotationHub

AnnotationHub provides a resource of annotations that are available without
requiring an annotation package.

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
```

```{r}
query(ah, "^org\\.")
```

```{r}
# ah["AH70563"]
# unique(ah$species)

# find out if contains string
# grepl(ah$species, "Homo")
# azolla <- subset(ah, species %in% "Homo")
# pmatch(ah$species, "azolla")
```

```{r}
org <- ah[["AH70563"]]
```

```{r}
AnnotationDbi::select(GO.id, gid, c("ENTREZID", "SYMBOL", "ENSEMBL"))
```
```

