---
title: "demo_5"
author: "Theo Killian"
date: "July 1, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GSEA

Either your gene is in the gene set or not. Differentially expressed or not. 

Enrichment score, sort genes bv log fold change, sort of a random walk

Self-contained null (the genes do not have any association with the subject
condition)

You want to permute subjects

The general answer is no, you shouldn't throw out non-significant genes

"tibbles don't believe in row names"

Goana compares your subset with the universe of all possible genes, and looks at
GO terms for "Hs" (human, for example), returns a tibble of the GO terms that
are significantly expressed

"Deep diving, bringing up shadowy gold"

## Theory

See [slides][]

<!-- [slides]: https:// -->

## Benchmarking: a recent tweet...

A recent [tweet][] provides a nice summary of efforts to benchmark
gene set enrichement analysis methods using the [GSEABenchmarkR][]
package.

```{r, message=FALSE}
library(EnrichmentBrowser)
```

# Practice

```{r, message = FALSE, echo = FALSE}
library(DESeq2)
library(airway)
library(dplyr)
library(org.Hs.eg.db)
library(GO.db)
library(limma)
```

Data input and massage

```{r}
library(airway)
data(airway)
airway$dex <- relevel(airway$dex, "untrt")
```

Differential expression analysis

```{r}
library(DESeq2)
des <- DESeqDataSet(airway, design = ~ cell + dex)
des <- DESeq(des)
res <- results(des)
```

Transition to tidy data

```{r}
library(dplyr)
library(tibble)
tbl <- res %>%
    as.data.frame() %>%
    rownames_to_column("ENSEMBL") %>%
    as_tibble()
tbl
```

## Example: hypergeometric test using [limma][]`::goana()`

Requires ENTREZ identifiers

```{r}
library(org.Hs.eg.db)
tbl <- tbl %>% 
    mutate(
        ENTREZID = mapIds(
            org.Hs.eg.db, ENSEMBL, "ENTREZID", "ENSEMBL"
        ) %>% unname()
    )
tbl
```

Universe -- must be testable for DE

```{r}
tbl <- tbl %>% 
    filter(!is.na(padj), !is.na(ENTREZID))
tbl
```

[limma][]`::goana()` -- Hypergeometric

```{r}
library(limma)
go <- 
    goana(tbl$ENTREZID[tbl$padj < .05], tbl$ENTREZID, "Hs") %>%
    as_tibble()
```

Hmm, `goana()` returns GO terms, but we also need GO identifiers

```{r}
library(GO.db)
go <- 
    go %>%
    mutate(
        GOID = mapIds(
            GO.db, .$Term, "GOID", "TERM"
        ) %>% unname()
    ) %>%
    select(GOID, everything()) %>% 
    arrange(P.DE)
```

Sanity check

```{r}
go %>% filter(grepl("glucocorticoid", Term))
```

What genes in set?

```{r}
genesets <- 
    AnnotationDbi::select(org.Hs.eg.db, tbl$ENTREZID, "GO", "ENTREZID") %>%
    as_tibble() %>%
    select(ENTREZID, GO, ONTOLOGY) %>%
    distinct()
genesets
```

# Provenance

```{r}
sessionInfo()
```

[limma]: https://bioconductor.org/packages/limma
[GSEABenchmarkR]: https://bioconductor.org/packages/GSEABenchmarkR
[tweet]: https://twitter.com/LeviWaldron1/status/1142092301403115521